XRONOS: an open data infrastructure for archaeological chronometry

### 2024-01-30 paper.qmd
Redo outline

---
title: "paper"
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

```{r}
1 + 1
```

>>>>

---
title: "XRONOS: an open data infrastructure for archaeological chronometry"
author:
  - name: Joe Roe
    orcid: 0000-0002-1011-1244
    url: "https://joeroe.io"
    email: joseph.roe@unibe.ch
    affiliation:
    - ref: iaw
    roles: [writing, editing, software, analysis]
  - name: Martin Hinz
    orcid: 0000-0002-9904-6548
    url: "https://martinhinz.info"
    email: martin.hinz@unibe.ch
    corresponding: true
    affiliation:
    - ref: iaw
    roles: [conceptualization, funding, supervision, editing, software]
affiliations:
  - id: iaw
    name: University of Bern
    department: Institute of Archaeological Sciences
    country: Switzerland
    url: https://www.iaw.unibe.ch/
abstract: |
  XRONOS (<https://xronos.ch>) is an open data infrastructure for the backbone of the archaeological record – chronology. It provides open access to published radiocarbon dates and other chronometric data from any period, anywhere in the world.

  Chronometric data has been one of the principle domains in which open data and open science approaches have been applied over the last ten years. Comprehensive compilations of radiocarbon dates have become available for many parts of the world in the last decade and, as natural next step, there are now several initiatives to collate this data globally, including the retrieval tool c14bazAAR [@SchmidEtAl2019], the IntChron exchange format [@BronkRamseyEtAl2019], and the synthetic database p3kc14 [@BirdEtAl2022]. This move from archiving individual datasets to producing a living open data infrastructure for archaeological chronologies raises a number of challenges. Radiocarbon datasets are still sorely lacking for many parts of the world and, even in those regions with good coverage, the quality of data is highly uneven and largely undocumented. There exists no central repository ensuring the long-term sustainability and completeness of these datasets, and the potential of placing other sources of chronometric information (e.g. dendrochronology, typological dating) in an open data framework has hardly been realised at all.

  Here, building on and complementing these initiatives, we present XRONOS: a new web-based platform for chronometric data from archaeological contexts worldwide, combining an open data repository with tools for importing, curating and analysing chronometric information from diverse sources. The principal design goals of the software are to combine all available sources of radiocarbon and other chronometric data in single database; develop robust tools for the continuous ingestion and refinement of this data; and disseminate this data within an open and FAIR framework, embedding it in the wider world of Linked Open Data in archaeology and beyond.

  This paper outlines the conceptual and technical infrastructure developed to realise these goals in XRONOS’ initial phases of development (2019 and 2021–2023), including a generalised data model for site and radiocarbon information, extendable to other chronometric data; an R- and Ruby-based pipeline for continuous ingestion of data from a variety of sources; continuous, semi-automated data cleaning protocols; a Ruby-on-Rails application providing a web-based frontend to the data and a REST API for programmatic access; and an R package for interfacing with the API. We believe the XRONOS framework provides more open, more reliable, and more comprehensive access to chronometric data than previously available, as well as a foundation for its continuous expansion and refinement.
copyright: 
  holder: the authors
  year: 2023
license: "CC BY"
# citation: 
#  container-title: Archaeometry
#  volume: 1
#  issue: 1
#  doi: 10.5555/12345678
funding: "This work was funded by the Swiss National Science Foundation ([SNSF Project #198152](https://data.snf.ch/grants/grant/198153))"
bibliography: references.bib
format:
  html:
    page-layout: article
    toc: true
---

## Introduction

* Chronometric modelling depends on ready access to chronometric data!
* Luckily, we do have a lot of it.
* Moving from 'open data' (uploading a CSV) to 'open infrastructure' [@CITE]

In this paper we:

* Review the state of the art in radiocarbon data compilation (and other chronometric data, such as they are), highlighting the need for XRONOS
* Describe XRONOS' software architecture, design goals, and currently-implemented features
  * With a focus on the choices we made in representing/modelling chronological data
* Assess the current state of the software and database, and give case studies of how it can be used
* Conclude with our future plans

## Compilations of radiocarbon dates

* Radiocarbon dates used to be regularly compiled and published in *Radiocarbon* (so it was 'open data' before it was cool), but this is no longer practical.
* Plus the bare measurements are of limited use without contextual information, which labs don't usually have access to.
* So even though radiocarbon data comes from a relatively limited number of sources [@IntCalLabList] and has relatively standardised reporting conventions [@Millard2014], in practice data has to be manually compiled from literature – this is time-consuming!
* Since at least 19XX, there have been efforts to systematically collect and compile this information
* Graph of number of dates/databases through time
* But this has been fragmentary: we counted at least XX databases
* And patchy: highly uneven spatial and geographic coverage, different data collected, different conventions for recording data

## Time for a global radiocarbon database?

* Regional databases are put together *ad hoc* and rarely continuously maintained over long periods
* The coverage of these databases is uneven, systematically biased, and duplicative of each other
* Other forms of chronometric data (dendro, typology, sea level, etc.) remain relatively untouched

* Recent moves towards a global synthesis: c14bazAAR [@SchmidEtAl2019], IntChron [@BronkRamseyEtAl2019], p3k14c [@BirdEtAl2022]
* Each taking a different approach, pros and cons
* We see a need for something that is comprehensive, unified, living
* And we don't see why it should be limited to radiocarbon: what about other radiometric dates? Dendro? Typologies?

## XRONOS

### Design goals and principles

* Inspired by GBIF, which manages to integrate large amounts of data from diverse sources into a signal, consistent database

### Data model

* Basic data structure inherited from RADON.
* Chrons: coordinates linking a time, a place, and human activity
  * Radiocarbon dates, typos, dendro etc.
  * Departs from the usual convention of treating e.g. a "period" as a different sort of information to a scientific date
  * In fact they can all be seen as the same kind of data
  * Time element can be represented as a probability distribution
  * Facilitates downstream modelling in e.g. OxCal, aoristic
* Contextual information: sample, context, site
* Bibliographic references
* Paper trail

### Software architecture

* CRUD Rails app
  * Not so exciting, but try to keep up with modern standards, remain accessible, etc.
  * Data browser
* API
  * R package
* Linked data

## Implementation

The first prototype of XRONOS was developed in 2019.
It has been under sustained development 2021–2024 with the first public release in July 2021.

* Assessment of XRONOS' coverage
* Tale from the trenches: duplicate processing?
* Tale from the trenches: cleaning species and countries?
* Case study of use?

## Conclusion

* Chronology is the backbone of the archaeological record
* XRONOS could be a backbone for a universe of linked open data resources...
* ...but sustainability is a challenge.
  * We're coming to the end of our project funds
  * We'll try to get more, but you never know (grant-based science and long-term investments in digital infrastructure don't work so well together!)
  * Future sustainability

## Acknowledgements

Direct contributors to XRONOS (or invite as coauthors?):

* Clemens Schmid
* Caroline Heitz
* Albert Hafner
* Setareh Ebrahimiabareghi
* Chiara Huwiler
* Rivana Moser
